---
title: "Causal Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Introduction

The package `causalClust` implements methods for designing cluster-randomized experiments under network interference. In this setting, one unit's treatment may affect neighboring units. The choice of experimental clusters therefore affects both statistical precision and spillover bias.

The package is designed for experiments where the user observes an experimental network and wants to choose clusters before randomizing treatment. The central input is an adjacency matrix `W`; the central output is a vector of cluster labels.

The main workflow is:

1. provide an adjacency matrix for the experimental network;
2. choose a calibration value `xi`, or provide a calibration grid/range;
3. run `causal_clustering_algorithm()`;
4. inspect the selected clustering, number of clusters, objective components, and optimization diagnostics;
5. compare the selected clustering with simpler baselines when useful.

## Reference

This package implements methods from:

Viviano, D., Lei, L., Imbens, G., Karrer, B., Schrijvers, O., & Shi, L. (2026). *Causal clustering: design of cluster experiments under network interference*. Manuscript, May 28, 2026.

## Installation

Install the development version from GitHub with:

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("ostasovskyi/causalClustering")
```

Load the package with:

```{r, eval = FALSE}
library(causalClust)
```

Some features require optional packages:

```{r, eval = FALSE}
install.packages(c("igraph", "ggplot2", "Matrix"))
```

The SDP engine requires `Matrix` and `sdpt3r`. If `sdpt3r` is not available from CRAN in your R setup, install it from GitHub:

```{r, eval = FALSE}
install.packages("remotes")
remotes::install_github("AdamRahman/sdpt3r")
```

## Example

We first illustrate the main estimator on a simulated network. The goal is to choose experimental clusters that trade off variance and spillover bias.

```{r, eval = FALSE}
library(causalClust)

sim <- simulate_network_data(
  parameters_graph = list(
    n = 60,
    type_graph = "geometric",
    neighb = 2
  )
)

W <- sim$W
```

The object `W` is the adjacency matrix used by the clustering algorithm. Nonzero entries indicate edges in the experimental network.

We now run the causal clustering algorithm for a single calibration value.

```{r, eval = FALSE}
fit <- causal_clustering_algorithm(
  W = W,
  xi = 1,
  min_k = 2,
  max_k = 12,
  engine = "sdp",
  methods = available_discretization_methods(),
  seed = 123
)
```

After fitting, we can inspect the selected design.

```{r, eval = FALSE}
fit$algorithm
fit$selected_k
fit$selected_method
fit$objective
fit$components
head(fit$clusters)
```

The entry `selected_k` is the number of clusters selected by the algorithm. The entry `selected_method` records which discretization method was selected after rounding the relaxation and evaluating the realized objective.

## Objective components

For a candidate clustering, the package evaluates the objective

```text
xi * variance + bias^2.
```

The variance component is the normalized sum of squared cluster sizes. The bias component is the average fraction of each unit's neighbors assigned to a different cluster.

The objective can also be evaluated directly for a user-supplied clustering.

```{r, eval = FALSE}
W_small <- matrix(
  c(
    0, 1, 1, 0,
    1, 0, 0, 1,
    1, 0, 0, 1,
    0, 1, 1, 0
  ),
  nrow = 4,
  byrow = TRUE
)

clusters <- c(1, 1, 2, 2)

components <- clustering_objective_components(W_small, clusters)
objective <- compute_causal_clustering_objective(
  W = W_small,
  xi = 1,
  clusters = clusters
)

components
objective
all.equal(objective, components$variance + components$bias^2)
```

## Calibration grids

The calibration parameter `xi` controls the relative weight on the variance component. Larger values of `xi` place more weight on variance; smaller values place more weight on spillover-bias reduction.

If the calibration is uncertain, pass a grid or range. The same public function automatically uses the endpoint-regret algorithm.

```{r, eval = FALSE}
fit_grid <- causal_clustering_algorithm(
  W = W,
  xi_grid = c(0.2, 0.5, 1, 2),
  min_k = 2,
  max_k = 12,
  engine = "sdp",
  methods = available_discretization_methods(),
  seed = 123
)

fit_grid$algorithm
fit_grid$xi_range
fit_grid$selected_k
fit_grid$selected_method
fit_grid$objective
fit_grid$rho_values
fit_grid$components
```

The package also accepts `calibration`, `calibration_grid`, and `calibration_range`. These are converted internally by

```text
xi = 1 / calibration.
```

## Discretization methods

The relaxation output must be rounded into a feasible clustering. The supported rounding methods are:

```{r, eval = FALSE}
available_discretization_methods()
```

They are currently:

```{r, eval = FALSE}
c(
  "kmeans",
  "hierarchical",
  "spectral_norm_kmeans",
  "spectral_unnorm_kmeans",
  "spectral_norm_hierarchical",
  "spectral_unnorm_hierarchical"
)
```

When several methods are supplied, the package tries all requested methods and keeps the candidate with the smallest realized objective.

## SDP diagnostics

With `engine = "sdp"`, fitted objects may contain an SDP lower bound and an approximation certificate.

```{r, eval = FALSE}
fit$Gamma_n
fit$certificate_valid
fit$sdp_lower_bound
```

The argument `k_constraint` controls how the SDP relaxation is solved.

```{r, eval = FALSE}
fit_k <- causal_clustering_algorithm(
  W = W,
  xi = 1,
  min_k = 2,
  max_k = 12,
  engine = "sdp",
  k_constraint = TRUE,
  gamma_bar = 10,
  box_constraints = TRUE,
  methods = available_discretization_methods(),
  seed = 123
)
```

With `k_constraint = FALSE`, the algorithm solves one relaxation and rounds it over the candidate K grid. With `k_constraint = TRUE`, it solves a K-specific SDP relaxation for each candidate K and can impose additional cluster-size and box constraints.

## Spectral approximation

The package can be used without an SDP solver by setting `engine = "spectral"`. This is useful for exploratory computation. The spectral engine is available for scalar calibration values.

```{r, eval = FALSE}
fit_spectral <- causal_clustering_algorithm(
  W = W,
  xi = 1,
  min_k = 2,
  max_k = 12,
  engine = "spectral",
  methods = available_discretization_methods(),
  seed = 123
)

fit_spectral$selected_k
fit_spectral$selected_method
fit_spectral$objective
fit_spectral$components
```

The spectral engine evaluates the same realized objective after rounding, but it does not provide the SDP lower bound or approximation certificate.

## Baseline clustering methods

The package also includes baseline clustering procedures.

```{r, eval = FALSE}
epsilon_fit <- cluster_epsilon_net(W)
spectral_fit <- cluster_spectral(W, num_clusters = 5, seed = 123)
```

If `igraph` is installed, Louvain clustering is also available.

```{r, eval = FALSE}
louvain_fit <- cluster_louvain_membership(W)
```

These baselines can be evaluated using the same objective functions.

```{r, eval = FALSE}
compute_causal_clustering_objective(W, xi = 1, clusters = epsilon_fit$clusters)
compute_causal_clustering_objective(W, xi = 1, clusters = spectral_fit)
compute_causal_clustering_objective(W, xi = 1, clusters = louvain_fit)
```

## Simulation comparison

The function `run_single_network()` simulates or accepts a network, runs the causal clustering method over a sequence of `xi` values, and compares it with baseline clusterings.

```{r, eval = FALSE}
out <- run_single_network(
  seed = 123,
  parameters_graph = list(
    type_graph = "geometric",
    n = 100,
    neighb = 2
  ),
  xi_seq = seq(0.1, 5, by = 0.5),
  min_k = 2,
  max_k = 20,
  include_louvain = TRUE,
  engine = "spectral",
  objective_type = "squared",
  methods = available_discretization_methods()
)

head(out$results)
```

The plotting functions summarize the objective path and the selected number of clusters.

```{r, eval = FALSE}
plots <- plot_objective_path(out$results)
plots$plot_objective
plots$plot_num_clusters
```

## Additional parameters

The most commonly changed parameters are:

```{r, eval = FALSE}
causal_clustering_algorithm(
  W = W,
  xi = 1,
  min_k = 2,
  max_k = 12,
  n_eig = NULL,
  engine = "sdp",
  methods = available_discretization_methods(),
  include_bernoulli = FALSE,
  objective_type = "squared",
  k_constraint = FALSE,
  gamma_bar = 10,
  box_constraints = TRUE,
  try_sign_flip = TRUE,
  keep_search_matrices = FALSE,
  keep_sdp_solutions = FALSE,
  seed = 123
)
```

For most users, the main choices are the adjacency matrix `W`, the calibration input, the K search range, and the optimization engine.
